canaper: Categorical analysis of neo- and paleo-endemism in R

### 2022-10-14 ms/ms.Qmd
Update MS
fix some problems with refs, move Fig S1 code to SI

\* Corresponding author  
Joel H. Nitta  
Department of Integrated Biosciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan  
joelnitta@gmail.com

Running headline: *canaper*

@@

It is useful to plot the output to identify the minimum number of iterations needed.
Here we do so with the `ggplot2` package.

```{r}
#| label: cpr-iter-sim-plot
#| fig.height: 8
#| eval: false

library(ggplot2)

ggplot(iter_sim_res, aes(x = iteration, y = similarity)) +
  geom_line() +
  labs(x = "Num. iterations", y = "% Similarity")
```

The plot shows that the community matrix becomes maximally randomized after about 40,000--50,000 iterations (@fig-iter-sim), and that the original matrix and randomized matrix are about 96.5% similar.

@@

a + b + plot_layout(ncol = 1) + plot_annotation(tag_levels = "a")
```

@@

The color palette used in @fig-acacia-cpr, `mishler_endem_cols`, is the same as that used in @Mishler2014.
However, this palette may not be distinguishable to people with color-vision deficiency (CVD).
Alternate color palettes (e.g., `cpr_endem_cols`) are also available in `canaper` that include CVD-safe colors [@Okabe2002].
Palettes for plotting CANAPE and *p*-rank results can be selected with the `cpr_make_pal()` function.
The same plot visualized with CVD-safe colors is available in the online supporting material (Fig. S1).

@@

<!-- FIXME: for double-blind review leave out of submission text, but add after accepted 
To cite canaper or acknowledge its use, cite this Software note as follows, substituting the version of the application that you used for 'version 1.0.0':

Nitta, J. H. et al. 2022. canaper: Categorical analysis of neo- and paleo-endemism in R. FIXME: ADD DETAILS HERE (ver 1.0.0) -->

<!-- FIXME: for double-blind review leave out of submission text, but add after accepted 
# Acknowledgements

Members of the Iwasaki Lab (The University of Tokyo) provided helpful comments while the package was in development.
Luis Osorio and Klaus Schliep reviewed the code as part of submission to rOpenSci.
-->

@@

Joel H. Nitta conceived the ideas and designed methodology; Joel H. Nitta and Shawn W. Laffan analyzed the data; Joel H. Nitta led the writing of the manuscript; Brent D. Mishler edited the manuscript; Wataru Iwasaki provided resources and supervised the study.
All authors contributed critically to the drafts and gave final approval for publication.

@@

# References
<!-- Insert <div id="refs"></div> to place the refs here instead of at the end -->
<div id="refs"></div>

@@

**@fig-iter-sim.** Testing the number of iterations needed to randomize the Australian *Acacia* dataset.
The *Acacia* dataset [@Mishler2014] was randomized using the `curveball` algorithm for 100,000 iterations and percentage similarity between the original matrix and the randomized matrix was calculated every 1,000 iterations using the `cpr_iter_sim()` function.

@@

```{r}
#| label: fig-iter-sim
#| fig.height: 4
#| fig-cap: "Testing the number of iterations needed to randomize the Australian *Acacia* dataset."
#| echo: false

fig <- ggplot2::ggplot(iter_sim_res, aes(x = iteration, y = similarity)) +
  geom_line() +
  labs(x = "Num. iterations", y = "% Similarity")

ggsave(plot = fig, file = here::here("ms/figure_2.pdf"),
  width = 150, height = 150, units = "mm")

@@

fig <- a + b + plot_layout(ncol = 1) + plot_annotation(tag_levels = "a")

ggsave(plot = fig, file = here::here("ms/figure_3.pdf"),
  width = 100, height = 200, units = "mm")

>>>>

\* Corresponding author  
Joel H. Nitta  
joelnitta@gmail.com

@@

It is useful to plot the output to identify the minimum number of iterations needed.
Code to generate a plot with the `ggplot2` package and its output are shown in the online Supporting Information (Fig. S1).
The plot shows that the community matrix becomes maximally randomized after about 40,000--50,000 iterations and that the original matrix and randomized matrix are about 96.5% similar.

@@

a + b + plot_layout(ncol = 1) + 
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")")
```

@@

The color palette used in @fig-acacia-cpr, `mishler_endem_cols`, is the same as that used in @Mishler2014.
However, this palette may not be distinguishable to people with color-vision deficiency (CVD).
Alternate color palettes (e.g., `cpr_endem_cols`) are also available in `canaper` that include CVD-safe colors [@Okabe2002].
Palettes for plotting CANAPE and *p*-rank results can be selected with the `cpr_make_pal()` function.
The same plot visualized with CVD-safe colors is available in Fig. S2.

@@

<!-- FIXME: for double-blind review leave out of submission text, but add after accepted

To cite canaper or acknowledge its use, cite this Software note as follows, substituting the version of the application that you used for 'version 1.0.0':

Nitta, J. H. et al. 2022. canaper: Categorical analysis of neo- and paleo-endemism in R. FIXME: ADD DETAILS HERE (ver 1.0.0)

# Acknowledgements

Members of the Iwasaki Lab (The University of Tokyo) provided helpful comments while the package was in development.
Luis Osorio and Klaus Schliep reviewed the code as part of submission to rOpenSci.

@@

**Joel H. Nitta**: Conceptualization (Lead); Formal analysis (Lead); Funding acquisition (Supporting); Methodology (Lead); Software (Lead); Visualization (Lead); Writing – original draft (Lead); Writing – review & editing (Equal).
**Shawn W. Laffan**: Data curation (Lead); Methodology (Supporting); Software (Supporting); Writing – review & editing (Equal).
**Brent D. Mishler**: Writing – review & editing (Equal).
**Wataru Iwasaki**: Funding acquisition (Lead); Project administration (Lead); Supervision (Lead); Writing – review & editing (Equal).

-->

@@

# References

<!-- FIXME: for some reason Lai2019 shows up as "in press" -->
<div id="refs"></div>

@@

@@

@@

fig <- a + b + plot_layout(ncol = 1) + plot_annotation(
  tag_levels = "a",
  tag_prefix = "(",
  tag_suffix = ")")

ggsave(plot = fig, file = here::here("ms/figure_2.pdf"),
  width = 100, height = 200, units = "mm")

### 2022-10-14 ms/si.Qmd
Update MS
fix some problems with refs, move Fig S1 code to SI

---
title: "canaper: Categorical analysis of neo- and paleo-endemism in R"
subtitle: "Supporting information"
author:
  - Joel H. Nitta
  - Shawn W. Laffan
  - Brent D. Mishler
  - Wataru Iwasaki
bibliography: references.yaml
csl: methods-in-ecology-and-evolution.csl
highlight-style: github
format:
  pdf:
    include-in-header:
      # keep figures from floating away
      text: |
        \usepackage{float}
        \makeatletter\renewcommand*{\fps@figure}{H}\makeatother
---

@@

@@

**Fig. S1**: Categorical analysis of neo- and paleo-endemism (CANAPE) of Australian *Acacia*, plotted with color vision deficiency (CVD)-safe colors.
(a) Map of Australia showing grid-cells (communities) colored by endemism type.
Latitude and longitude projected into equal area Australian Albers (EPSG:3577) coordinate system; units in m.
(b) Scatterplot of comparing phylogenetic endemism (PE) of each community as measured on the original tree vs. a comparison tree with all non-zero branch lengths set to equal length, colored according to endemism type.

>>>>

---
title: "canaper: Categorical analysis of neo- and paleo-endemism in R"
subtitle: "Supporting information"
# FIXME: for double-blind review leave out of submission text, but add after accepted 
# author:
#   - Joel H. Nitta
#   - Shawn W. Laffan
#   - Brent D. Mishler
#   - Wataru Iwasaki
bibliography: references.yaml
csl: methods-in-ecology-and-evolution.csl
highlight-style: github
format:
  pdf:
    include-in-header:
      # keep figures from floating away
      text: |
        \usepackage{float}
        \makeatletter\renewcommand*{\fps@figure}{H}\makeatother
---

@@

```{r}
#| label: iter-sim
#| fig.height: 4
#| echo: false

set.seed(12345)

data(acacia)

# Conduct up to 100,000 iterations (swaps),
# recording similarity every 1000 iterations
iter_sim_res <- cpr_iter_sim(
  comm = acacia$comm,
  null_model = "curveball",
  n_iterations = 100000,
  thin = 1000
)

fig <- ggplot2::ggplot(iter_sim_res, aes(x = iteration, y = similarity)) +
  geom_line() +
  labs(x = "Num. iterations", y = "% Similarity")

fig
```

**Fig S1.** Testing the number of iterations needed to randomize the Australian *Acacia* dataset.
The *Acacia* dataset was randomized using the `curveball` algorithm for 100,000 iterations and percentage similarity between the original matrix and the randomized matrix was calculated every 1,000 iterations using the `cpr_iter_sim()` function.

Code to generate Fig. S1 is as follows.

```{r}
#| label: cpr-iter-sim-show
#| eval: false
library(canaper)
library(ggplot2)

set.seed(12345)

iter_sim_res <- cpr_iter_sim(
  comm = acacia$comm,
  null_model = "curveball",
  n_iterations = 100000,
  thin = 1000
)

ggplot(iter_sim_res, aes(x = iteration, y = similarity)) +
  geom_line() +
  labs(x = "Num. iterations", y = "% Similarity")
```

@@

**Fig. S2**: Categorical analysis of neo- and paleo-endemism (CANAPE) of Australian *Acacia*, plotted with color vision deficiency (CVD)-safe colors (`cpr_endem_cols_4` palette in `canaper`).
(a) Map of Australia showing grid-cells (communities) colored by endemism type.
Latitude and longitude projected into equal area Australian Albers (EPSG:3577) coordinate system; units in m.
(b) Scatterplot of comparing phylogenetic endemism (PE) of each community as measured on the original tree vs. a comparison tree with all non-zero branch lengths set to equal length, colored according to endemism type.
